Actual Causes as Changes of State in Continuous Time

### 2021-01-20 docs/resources/CogSci_paper/paper.tex
edited expt 1

We developed two experiments to test the continuity account of actual causation just presented.  The first experiment explores some simple cases directly analogous to the smoke detector example. The second experiment explores cases that elaborate that distinguish between the continuity account and Spellman's probability raising account~\cite{spellman97}.

The cover story for both experiments asked participants to imagine that they worked in a control room and that their job was to monitor networks of particle detectors. Detectors were shown as white squares (as in Figure~\ref{fig:intro} that become active and turn black when they absorb a radioactive particle. Activation is transmitted across links in the network, and a detector activates if all of its input detectors are active. Each network contains a special detector called the Gauge of Critical Moment (GCM, analogous to the smoke alarm in Figure~\ref{fig:intro}a) and participants were told that ``for each activation sequence that you see, your job is to decide what caused the activation of the GCM.'' The prompt after each sequence was ``In this sequence what caused the activation of the GCM? Respond by clicking on a detector,'' and participants were required to choose a single detector in response.

Both experiments included chains in which the GCM received input from one detector, and dual branch networks such as Figure~\ref{fig:intro}a in which the GCM received input from two detectors. For chains, the continuity account predicts that participants will tend to choose the root cause of the effect (ie the detector that initiates the activation sequence) rather than the immediate cause (the detector that immediately precedes the GCM). For dual branch networks, the continuity account predicts that participants will tend to choose the root cause on the branch whose activation is temporally continuous with the activation of the GCM. We refer to this branch as the continuous branch, and refer to the root cause and the immediate cause on this branch as \ev{R}{C} and \ev{I}{C} respectively, where the subscript denotes ``continuous.'' We refer to the other branch as the ``delayed branch,'' and use \ev{R}{D} and \ev{I}{D} for the root and immediate cause on this branch. The ``delay'' in this naming scheme refers to the delay that occurs between the activation of \ev{I}{D} and the activation of the GCM.

@@

The purpose

The objective of this experiment was to see whether people's causal judgement were influenced by: the length of the sequences of changes (for chains and AND-Gates), and -- for AND-Gates only -- the delay between the last activation of the first branch (I1) and the first activation of the second branch (R2). In this experiment activation starts to spread over the second branch after the activation reached the end of the first branch (like in Fig.\ref{fig:1}).

@@

\textbf{Method}.

% Examples of network activations were shown in the instructions and participants had to answer to answer a survey at the end to see if they correctly understood the instructions. If they made a mistake they had to read again all the instructions and answer again the survey.

Each participant was presented 15 different networks of square detectors. Square detectors maintain their activation through time. Participants were first presented the network in its initial and static state and were asked to click the ``Run'' button to observe an activation sequence over the network. They waited 5000ms to see a first change of state occurring in a detector and the activation delay between any two successive detectors was set on 100ms.

Three stimuli were chains that differed in length of activation sequence, that is including 2, 4 or 7 activated squares (with the GCM). These lengths were labeled ``Short'', ``Medium'' and ``Long''. The AND-Gates were similarily categorised ``Short'', ``Medium'' (the one in Fig.\ref{fig:1}) and ``Long'' with both branches being the same length. In each length category participants were shown three similar AND-Gates that differed uniquely in the delays between I1 and R2. The delays were 2s, 4s and 6s.

All the stimuli were presented in a random order. Two groups, ``Left'' and ``Right'', were designed and partcipants were randomly assigned to either group at the begining of the experiment. In the ``Left'' group all the networks were shifted towards left (the GCM being on the left) and in the ``Right'' group all the networks were shifted towards right. For each AND-Gate participants saw randomly either the top branch or the bottom branch activating first.

After the activation of the GCM (the end of the animation) participants had to wait 1000ms before the squares became clickable and the following intruction appeared: ``In this sequence what caused the activation of the GCM? Respond by clicking on a detector''. They had the option to run again the animation (no more than 9 times) or to go to the next network assuming they clicked on one detector -- they couldn't select more than one detector. When selected the edges of a detector turned red.

\textbf{Results}. First of all we observed no significant effect [STATISTICS?] of neither the orientation of the networks (left \textit{vs} right) nor the location where the activation starts first (top \textit{vs} bottom). On average participants took 14 min to complete the experiment and overall 96\% of the time participants run an animation only once. Second of all the results show two main findings: first the length of the activation sequences (one, three or six detectors in each branch of the AND-Gates) has no effect on the response pattern; second neither the delay between I1 and R2 (2000ms, 4000ms and 6000ms) nor the fact that some detectors were already activated in State condition have a significant effect on causal judgement. So overall people massively chose the last root change R2 as the main cause of the GCM activation for medium and long networks (and R2/I2 for the short networks since both were identical). For the short networks (state and event types combined) 25\% \textit{vs} 65\% of participants chose respectively R1/I1 \textit{vs} R2/I2 as the main cause; for the medium networks (state and event combined) we got 23\% \textit{vs} 53\% for respectively R1 and R2 and 20\% \textit{vs} 55\% for the long networks (Fig.\ref{fig:2}).

[WARNING: THE GRAPH BELOW SHOW THE RESULTS FOR SHORT, MEDIUM, LONG \textbf{LENGTHS} (ALL THE DELAYS MIXED) SO NOT THE DIFFERENT DELAYS BETWEEN I1 AND R2. BUT AS OUR FOCUS IS ON TEMPORAL INFORMATION, MAYBE WE SHOULD GRAPHICALLY SHOW THAT THE DELAY HAS NO EFFECT?? BUT IT MEANS THAT WE HAVE TO ADD TWO MORE COLUMNS FOR EACH DETECTOR CATEGORY]

\begin{figure}[ht]
\begin{center}
%\includegraphics[width=8cm]{results_E1}
\includegraphics{figures/expt1.eps}
\end{center}
\caption{Comparison of the results for different length of activation in AND-Gates. Short: 1 activated detector in each branch. Medium: 3 activated detectors in each branch. Long: 6 activated detectors in each branch. ``State'' responses correspond to the networks where one branch was already activated before participants run the animation. ``Event'' responses correspond to the networks where no detectors were activated before and include results for 2s, 4s and 6s delays between I1 and R2. BUT SEE WARNING ABOVE.} 
\label{fig:2}
\end{figure}

>>>>

We developed two experiments to test the continuity account of actual causation just presented.  
%The first experiment explores some simple cases directly analogous to the smoke detector example. The second experiment explores cases that elaborate that distinguish between the continuity account and Spellman's probability raising account~\cite{spellman97}.
The cover story for both experiments asked participants to imagine that they worked in a control room and that their job was to monitor networks of particle detectors. Detectors were shown as white squares (as in Figure~\ref{fig:intro} that become active and turn black when they absorb a radioactive particle. Activation is transmitted across links in the network, and a detector activates if all of its input detectors are active. Each network contains a special detector called the Gauge of Critical Moment (GCM, analogous to the smoke alarm in Figure~\ref{fig:intro}a) and participants were told that ``for each activation sequence that you see, your job is to decide what caused the activation of the GCM.''

Both experiments included chains in which the GCM received input from one detector, and dual branch networks such as Figure~\ref{fig:intro}a in which the GCM received input from two detectors. For chains, the continuity account predicts that participants will tend to choose the root cause of the effect (ie the detector that initiates the activation sequence) rather than the immediate cause (the detector that immediately precedes the GCM). For dual branch networks, the continuity account predicts that participants will tend to choose the root cause on the branch whose activation is temporally continuous with the activation of the GCM. We refer to this branch as the continuous branch, and refer to the root cause and the immediate cause on this branch as \ev{R}{C} and \ev{I}{C} respectively, where the subscript denotes ``continuous.'' We refer to the other branch as the ``delayed branch,'' and use \ev{R}{D} and \ev{I}{D} for the root and immediate causes on this branch. The ``delay'' in this naming scheme refers to the delay that occurs between \ev{I}{D} and the activation of the GCM. It is convenient to use the same labels for both detectors and events: for example, \ev{R}{C} will be used to denote both a detector on the continuous branch and the activation of this detector. This flexible use of notation, however, should not be taken to suggest that actual causation involves a relationship between objects instead of a relationship between events.

Our presentation of the experiments focuses on two theoretical accounts: the continuity account and the probability raising account. For completeness, though, we first discuss the counterfactual and physical process accounts. Both accounts can be formulated in different ways --- here we present what we take to be the most natural predictions of the two accounts, and the General Discussion considers ways in which the accounts can be adjusted to make different predictions.

The counterfactual account makes no clear prediction about whether root causes or immediate causes should take priority. For a chain network, if the root cause had not occurred, the effect would not have occurred, but it is equally true that if the immediate cause had not occurred, the effect would not have occurred. If we consider dual branch networks and restrict attention to root causes only, then the account makes no clear preediction about whether \ev{R}{C} should be preferred to \ev{R}{D}.  If \ev{R}{C} had not occurred, then the effect would not have occurred, and likewise for \ev{R}{D}.

The physical process account is similarly inconclusive. Both root and immediate causes are connected by a physical process to the effect, and there seem to be no clear grounds for preferring one to the another. If we focus only on root causes, again there is no clear preference between \ev{R}{C} and \ev{R}{D}. After the activation of \ev{I}{D}, one can think of this detector as continuously sending activation towards the GCM, but this unbroken stream of activation only ``unlocks`` the detector once \ev{I}{C} is also active. It follows that both \ev{R}{C} and \ev{R}{D} are connected to the effect by physical processes.

Because both of these accounts make no clear predictions, it seemed possible that people's inferences about the networks in our experiments would be highly variable and would reveal no clear trends. Our first experiment therefore explored simple cases analogous to the smoke detector example in Figure~\ref{fig:intro} with the goal of establishing whether the basic experimental procedure was viable.  The second experiment focused on more elaborate dual-branch cases that aimed to distinguish between the continuity account and Spellman's probability raising account~\cite{spellman97}.

@@

Experiment 1 included both causal chains and dual branch networks. For all of the dual branch stimuli, \ev{I}{D} occurred before \ev{R}{C} (Figure~\ref{fig:expt1}a) which means that the delayed branch completed its activation before the continuous branch began to activate. The probability raising and continuity accounts both predict that root causes should be preferred to immediate causes, and that for dual branch networks \ev{R}{C} should be preferred to \ev{R}{D}. The primary purpose of the experiment was to test both predictions.

@@

\textbf{Materials}. The experiment used a customized interface built using the jsPsych library~\cite{XXX}. For all networks presented, participants clicked a ``Run'' button to observe an activation sequence. The first event in the sequence (i.e.\ the first change of state) always took place 5 seconds after the Run button was clicked, and the delay between successive activations along a chain of detectors was set to 100 ms. After the final event in a sequence (i.e.\ the activation of the GCM), all detectors became clickable after a delay of 1 second. Clicking on a detector turned its border red, and at most one detector could be selected at any time. After a sequence completed, participants could view it again if they wished by clicking a ``Run again'' button.

\textbf{Design}.  The experiment included activation sequences over 15 networks (3 chains and 12 dual branch networks). Excluding the GCM, each chain and each branch of each dual network had 7 detectors. We refer to the sequences as \emph{short}, \emph{medium} or \emph{long} based on the distance between the root causes and the GCM. Excluding the GCM, the short, medium and long chain sequences showed 1, 3 and 6 active detectors respectively at the end of the sequence. Within each dual branch sequence, the root causes on the two branches (\ev{R}{D} and \ev{R}{C}) were equidistant from the GCM, but these distances varied across sequences. The sequence in Figure~\ref{fig:intro}a is a medium dual branch sequence (3 active detectors in each branch excluding the GCM), and short and long versions had 1 and 6 active detectors respectively per branch.

Each dual branch sequence (short, medium and long) came in four versions. The \emph{state} version showed the delayed branch (including detectors \ev{R}{D} \ev{I}{D}) as active from the very beginning of the sequence. The activation of this branch was therefore presented as having occurred at some indefinite time in the past, resulting in a steady state of activation. The three \emph{event} versions all showed \ev{R}{D} activating 5 seconds into the sequence, and had delays of 2, 4 and 6 seconds between the activation of \ev{I}{D} and \ev{R}{C}.

\textbf{Procedure}.

Participants first read a set of instructions which introduced the task and included examples of a chain and a dual branch network. They then answered three questions about the task and the detectors, and were sent back to read the instructions again if they answered incorrectly. They continued cycling through the instructions and the test questions until they answered all questions correctly.

The 15 activation sequences in the experiment proper were presented in random order. For dual branch sequences, the position of the delayed branch was also randomized (Figure~\ref{fig:intro} shows the delayed branch on the top rather than the bottom). The orientation of the network (GCM on the left or the right) was randomized within participants.  The prompt after each sequence was ``In this sequence what caused the activation of the GCM? Respond by clicking on a detector,'' and participants were required to choose a single detector in response.

\textbf{Results}.

Network orientation (left or right) and position of the delay branch (top or bottom) had no apparent effect, and we therefore collapse across these variables. Figure~\ref{fig:expt1} summarizes the results for short, medium and long dual-branch sequences. The delay between \ev{I}{D} and \ev{R}{C} had no significant effect, and Figure~\ref{fig:expt1} combines results for all three delays.

Although \ev{R}{D}, \ev{I}{D}, \ev{R}{C} and \ev{I}{C} all qualify as causes of the effect, Figure~\ref{fig:expt1} suggests that \ev{R}{C} tends to be singled out as the main cause. This result can be separated into two general conclusions.  First,  participants were more likely to choose root causes than immediate causes. Root and immediate causes were identical for the short sequences and therefore cannot bet distinguished, but the results for medium and long sequences reveal a preference for root causes. STATISTICAL ANALYSIS. The data for chains (not shown in Figure~\ref{fig:expt1}) provide further evidence or a preference for root causes. NUMBERS HERE

The second conclusion is that for dual branch sequences, the root cause on the continuous branch (\ev{R}{C}) is preferred to the root cause on the delayed branch (\ev{R}{D}). This result is consistent with the continuity account but also consistent with the probability raising account, because \ev{R}{C} guarantees the occurrence of the effect but \ev{R}{D} does not.

Figure \ref{fig:intro} suggests that event and state sequences lead to similar patterns of responses. The continuity account makes the same prediction for both kinds of sequences, but we thought that the event sequences might make participants especially likely to choose \ev{R}{C} over \ev{R}{D}. If this effect exists, our data suggest that it is relatively small.

It is also notable that people's judgments are largely unaffected by the delay between \ev{I}{D} and \ev{R}{C} and the activation length of each sequence (short, medium or long). Both of these variables affect the time that elapses between the root causes and the effect, and our results suggest that people's judgments are not exquisitely sensitive to this kind of variation.

%First of all we observed no significant effect [STATISTICS?] of neither the orientation of the networks (left \textit{vs} right) nor the location where the activation starts first (top \textit{vs} bottom). On average participants took 14 min to complete the experiment and overall 96\% of the time participants run an animation only once. Second of all the results show two main findings: first the length of the activation sequences (one, three or six detectors in each branch of the AND-Gates) has no effect on the response pattern; second neither the delay between I1 and R2 (2000ms, 4000ms and 6000ms) nor the fact that some detectors were already activated in State condition have a significant effect on causal judgement. So overall people massively chose the last root change R2 as the main cause of the GCM activation for medium and long networks (and R2/I2 for the short networks since both were identical). For the short networks (state and event types combined) 25\% \textit{vs} 65\% of participants chose respectively R1/I1 \textit{vs} R2/I2 as the main cause; for the medium networks (state and event combined) we got 23\% \textit{vs} 53\% for respectively R1 and R2 and 20\% \textit{vs} 55\% for the long networks (Fig.\ref{fig:2}).

%[WARNING: THE GRAPH BELOW SHOW THE RESULTS FOR SHORT, MEDIUM, LONG \textbf{LENGTHS} (ALL THE DELAYS MIXED) SO NOT THE DIFFERENT DELAYS BETWEEN I1 AND R2. BUT AS OUR FOCUS IS ON TEMPORAL INFORMATION, MAYBE WE SHOULD GRAPHICALLY SHOW THAT THE DELAY HAS NO EFFECT?? BUT IT MEANS THAT WE HAVE TO ADD TWO MORE COLUMNS FOR EACH DETECTOR CATEGORY]

\begin{figure}[ht]
\begin{center}
%\includegraphics[width=8cm]{results_E1}
\includegraphics{figures/expt1.eps}
\end{center}
\caption{Experiment 1. Comparison of the results for different length of activation in AND-Gates. Short: 1 activated detector in each branch. Medium: 3 activated detectors in each branch. Long: 6 activated detectors in each branch. ``State'' responses correspond to the networks where one branch was already activated before participants run the animation. ``Event'' responses correspond to the networks where no detectors were activated before and include results for 2s, 4s and 6s delays between I1 and R2. BUT SEE WARNING ABOVE.} 
\label{fig:2}
\end{figure}
